Tail associations and their impact on extinction risk

### 2019-11-05 MainText.Rmd
deleting all \nameref{} link

<!--ECEs probably produce tail associations-->
There are reasons to hypothesize that ECEs may produce asymmetric tail 
associations between climatic variables
measured in different locations. For instance, if a single ECE is associated
with spatially widespread, large values 
of some climatic variable, but small values of the
variable are associated with local weather rather than ECEs, the result could
be stronger right-tail association, through time, between measurements of the climatic variable
made in different locations.
For example, rainstorms which deposit exceptionally large amounts of rain
may do so over large areas; but periods with less rain may experience
scattered showers only, which may be much less correlated across locales.

@@

<!--stochastic matrix model : age-structured-->
To investigate Q1 from the \nameref{Introduction},
we considered a simple stochastic age-structured model
with two age classes, 
\begin{equation}
\left({\begin{array}{cc} E(t+1) \\ A(t+1) \end{array}}\right) = 
\left(\begin{array}{cc} 0 & f_a(t)\\ s_e(t) & 0 \end{array}\right)
\left(\begin{array}{c} E(t) \\ A(t) \end{array}\right), \label{eq_age_str}
\end{equation}
\noindent where $E(t)$ and $A(t)$ are the numbers or population densities of eggs 
and adults, respectively, at time $t$. We chose a very simple model
because our goal was to carry out an initial evaluation, only, of 
the realism of a potential mechanism. The stochastic vital rates $s_e(t)$ and $f_a(t)$
were selected, independently across times, as follows. 
The egg survival rate $s_e(t)$ was drawn from a beta distribution with parameters
$\alpha=3$ and $\beta=2$. 
Adult fecundity was $f_a(t)=\tilde{f}_a(t)\exp(-A(t)/\Omega)$, where $\tilde{f}_a(t)$
was drawn from a gamma distribution with shape and scale parameters $k=2$
and $\theta=1$, respectively, 
and $\Omega$ is a density dependence parameter. Taking $\Omega=\infty$ made
$\exp(-A(t)/\Omega)=1$ so that the model did not include density dependence
(it was then a standard stochastic matrix model). Finite, positive values of 
$\Omega$ made the model nonlinear, with density dependence in the adult
fecundity rate.

@@

<!--density-dependence metapopulation model-->
To investigate Q2 from the \nameref{Introduction}, we considered spatialized, stochastic versions
of six density dependent population models, the Ricker, Hassell, Maynard Smith, Pennycuick, Verhulst, 
and Austin-Brewer models.
All models took the general form $\vec{P} (t+1) = D \lambda(t) \vec{P}(t)$,
where $\vec{P}(t)$ is a vector of population densities, 
the $i^{\text{th}}$ component $P_i(t)$ representing
the population density in the $i^{\text{th}}$ habitat patch ($i=1,2,\ldots,n$) at time $t$. 
The $n \times n$ matrix $\lambda(t)$ was diagonal with $i^{\text{th}}$ diagonal entry 
$\lambda_i(t)$ depending on the model used (Table \ref{tab_model_summary}). The matrix $D$
was an $n \times n$ dispersal matrix. For simplicity, 
we assumed the $n$ habitat patches were arranged, evenly spaced, in a line.
We considered both "local" and "global" dispersal among patches in different simulations. 
For local dispersal, a fraction $d$ of 
each population dispersed during each time step, equally distributed to
the two or one nearest neighbors. For global dispersal the same fraction $d$ dispersed
equally to the other $n-1$ patches. When $d$ was $0$, dispersal did not occur and 
rescue effects were not possible.

@@

<!--test hypothesis from Q2 : ext risk with density-dependence metapopulation model with r from different regime-->
To answer (Q2) of the \nameref{Introduction}, 
we estimated extinction risk for each metapopulation model, using each form of tail 
association for the environmental noise,
using values of $r$ that included under- and overcompensatory dynamics,
as indicated above. We chose values of $r$ within a range, the
range specified as follows. The second part of Q2 (about how density dependence mediates the
influence of asymmetric tail associations on extinction risk) was addressed by comparing
results for under- and overcompensatory model regimes.
Defining $r_{min}$ as the minimum value of $r$ 
for which $P_e$ was non-negative; and defining
$r_{bf}$ as the value of $r$ at which a bifurcation occurs, the range of
$r$ used spanned 95\% of the interval ($r_{min},r_{bf}$). For details, see Appendix 
\ref{SM-model_stability}. The range of $r$ encompassed $r_c$, so included 
a transition from under- to overcompensatory dynamics. For each value of $r$
used within the range, we calculated extinction risk as indicated above.
Dispersal parameter values $d= 0,0.1,0.2,..0.5$ were also used in different 
simulations. We plotted extinction risk as a function of $d$ for two
fixed values of $r$, one producing under- and one overcompensatory dynamics.
For $d=0$ we also plotted extinction risk as a function of $r$. All code for 
this project is available at https://github.com/sghosh89/ERC.

@@

<!--Q2 : metapopulation results-->
For our metapopulation simulations, for low values of $r$, left-tail
associations of environmental noises accentuated extinction risks, and right-tail associations
mitigated them; but the reverse was true for high values of $r$ in the range
$(r_{min},r_{bf})$. Thus 
density dependent models can either produce similar results, regarding the influence of 
tail associations on extinction risk, as the density-independent models of @Ghosh_copula, or
opposite results, depending on the nature of density dependence. 
For each model of Table \ref{tab_model_summary}, using a value of $r$ on
the low end of the range $(r_{min},r_{bf})$, extinction risk was highest
for extreme left-tail associated noise, lowest for extreme right-tail associated
noise, and was intermediate for the moderate tail association scenarios, for
both local (Fig. \ref{fig_extrisk_allmodel_localdisp}, left panels) and global 
(Fig. \ref{SM-fig_extrisk_allmodel_globaldisp}, left panels) dispersal, and for 
essentially all the dispersal rates, $d$, that we considered. However, 
using a value of $r$ on the high end of the range $(r_{min},r_{bf})$,
extinction risks were ordered oppositely (right panels of Figs 
\ref{fig_extrisk_allmodel_localdisp} and \ref{SM-fig_extrisk_allmodel_globaldisp}). 
These results help address Q2 from the \nameref{Introduction}.
Intermediate values of $r$ are shown in Figs. \ref{SM-fig_ricker_risk_local_disp_scl_1} - 
\ref{SM-fig_risk_local_disp_abrewer_r_vary_scl_0.6} for local dispersal and 
Figs. \ref{SM-fig_ricker_risk_global_disp_scl_1} - 
\ref{SM-fig_risk_global_disp_abrewer_r_vary_scl_0.6} for global dispersal among 
habitat patches.

>>>>

<!--ECEs probably produce tail associations-->
There are reasons to hypothesize that ECEs may produce asymmetric tail 
associations between climatic variables
measured in different locations. For instance, suppose a single ECE is associated
with spatially widespread, large values 
of some climatic variable. Also suppose that small values of the
variable are associated, instead, with local weather, which can differ from location
to location, rather than with ECEs. The result could
be stronger right-tail association, through time, between measurements of the climatic variable
made in different locations.
For example, rainstorms which deposit exceptionally large amounts of rain
may do so over large areas. But periods with less rain may experience
scattered showers only, which may be much less correlated across locales.

@@

<!--stochastic matrix model : age-structured-->
To investigate Q1 from the Introduction,
we considered a simple stochastic age-structured model
with two age classes, 
\begin{equation}
\left({\begin{array}{cc} E(t+1) \\ A(t+1) \end{array}}\right) = 
\left(\begin{array}{cc} 0 & f_a(t)\\ s_e(t) & 0 \end{array}\right)
\left(\begin{array}{c} E(t) \\ A(t) \end{array}\right), \label{eq_age_str}
\end{equation}
\noindent where $E(t)$ and $A(t)$ are the numbers or population densities of eggs 
and adults, respectively, at time $t$. We chose a very simple model
because our goal was to carry out an initial evaluation, only, of 
the realism of a potential mechanism. The stochastic vital rates $s_e(t)$ and $f_a(t)$
were selected, independently across times, as follows. 
The egg survival rate $s_e(t)$ was drawn from a beta distribution with parameters
$\alpha=3$ and $\beta=2$. 
Adult fecundity was $f_a(t)=\tilde{f}_a(t)\exp(-A(t)/\Omega)$, where $\tilde{f}_a(t)$
was drawn from a gamma distribution with shape and scale parameters $k=2$
and $\theta=1$, respectively, 
and $\Omega$ is a density dependence parameter. Taking $\Omega=\infty$ made
$\exp(-A(t)/\Omega)=1$ so that the model did not include density dependence
(it was then a standard stochastic matrix model). Finite, positive values of 
$\Omega$ made the model nonlinear, with density dependence in the adult
fecundity rate.

@@

<!--density-dependence metapopulation model-->
To investigate Q2 from the Introduction, we considered spatialized, stochastic versions
of six density dependent population models, the Ricker, Hassell, Maynard Smith, Pennycuick, Verhulst, 
and Austin-Brewer models.
All models took the general form $\vec{P} (t+1) = D \lambda(t) \vec{P}(t)$,
where $\vec{P}(t)$ is a vector of population densities, 
the $i^{\text{th}}$ component $P_i(t)$ representing
the population density in the $i^{\text{th}}$ habitat patch ($i=1,2,\ldots,n$) at time $t$. 
The $n \times n$ matrix $\lambda(t)$ was diagonal with $i^{\text{th}}$ diagonal entry 
$\lambda_i(t)$ depending on the model used (Table \ref{tab_model_summary}). The matrix $D$
was an $n \times n$ dispersal matrix. For simplicity, 
we assumed the $n$ habitat patches were arranged, evenly spaced, in a line.
We considered both "local" and "global" dispersal among patches in different simulations. 
For local dispersal, a fraction $d$ of 
each population dispersed during each time step, equally distributed to
the two or one nearest neighbors. For global dispersal the same fraction $d$ dispersed
equally to the other $n-1$ patches. When $d$ was $0$, dispersal did not occur and 
rescue effects were not possible.

@@

<!--test hypothesis from Q2 : ext risk with density-dependence metapopulation model with r from different regime-->
To answer (Q2) of the Introduction, 
we estimated extinction risk for each metapopulation model, using each form of tail 
association for the environmental noise,
using values of $r$ that included under- and overcompensatory dynamics,
as indicated above. We chose values of $r$ within a range, the
range specified as follows. The second part of Q2 (about how density dependence mediates the
influence of asymmetric tail associations on extinction risk) was addressed by comparing
results for under- and overcompensatory model regimes.
Defining $r_{min}$ as the minimum value of $r$ 
for which $P_e$ was non-negative; and defining
$r_{bf}$ as the value of $r$ at which a bifurcation occurs, the range of
$r$ used spanned 95\% of the interval ($r_{min},r_{bf}$). For details, see Appendix 
\ref{SM-model_stability}. The range of $r$ encompassed $r_c$, so included 
a transition from under- to overcompensatory dynamics. For each value of $r$
used within the range, we calculated extinction risk as indicated above.
Dispersal parameter values $d= 0,0.1,0.2,..0.5$ were also used in different 
simulations. We plotted extinction risk as a function of $d$ for two
fixed values of $r$, one producing under- and one overcompensatory dynamics.
For $d=0$ we also plotted extinction risk as a function of $r$. All code for 
this project is available at https://github.com/sghosh89/ERC.

@@

<!--Q2 : metapopulation results-->
For our metapopulation simulations, for low values of $r$, left-tail
associations of environmental noises accentuated extinction risks, and right-tail associations
mitigated them; but the reverse was true for high values of $r$ in the range
$(r_{min},r_{bf})$. Thus 
density dependent models can either produce similar results, regarding the influence of 
tail associations on extinction risk, as the density-independent models of @Ghosh_copula, or
opposite results, depending on the nature of density dependence. 
For each model of Table \ref{tab_model_summary}, using a value of $r$ on
the low end of the range $(r_{min},r_{bf})$, extinction risk was highest
for extreme left-tail associated noise, lowest for extreme right-tail associated
noise, and was intermediate for the moderate tail association scenarios, for
both local (Fig. \ref{fig_extrisk_allmodel_localdisp}, left panels) and global 
(Fig. \ref{SM-fig_extrisk_allmodel_globaldisp}, left panels) dispersal, and for 
essentially all the dispersal rates, $d$, that we considered. However, 
using a value of $r$ on the high end of the range $(r_{min},r_{bf})$,
extinction risks were ordered oppositely (right panels of Figs 
\ref{fig_extrisk_allmodel_localdisp} and \ref{SM-fig_extrisk_allmodel_globaldisp}). 
These results help address Q2 from the Introduction.
Intermediate values of $r$ are shown in Figs. \ref{SM-fig_ricker_risk_local_disp_scl_1} - 
\ref{SM-fig_risk_local_disp_abrewer_r_vary_scl_0.6} for local dispersal and 
Figs. \ref{SM-fig_ricker_risk_global_disp_scl_1} - 
\ref{SM-fig_risk_global_disp_abrewer_r_vary_scl_0.6} for global dispersal among 
habitat patches.
